---
title: "Using the STAC catalogue to load current and future climate data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("gdalcubes")
library("rstac")
library("tibble")
library("sp")
library("sf")
library("dplyr")
library("rgbif")
library("tidyr")
library("stars")
```


```{r cars, echo=FALSE}
source("stac_functions.R")
```

### Exploring the STAC

We print the different collections from the STAC:

* chelsa-clim: current climate data (mean from 1981 to 2011) from CHELSA, categories "bioclim" (see the definitions of variables [here](https://www.worldclim.org/data/bioclim.html))
* chelsa-clim-proj: climate data (bioclim list) projected from 2041-2070 and 2071-2100, for different climatic scenario and CHELSA climatic model (see [CHELSA project](https://chelsa-climate.org/cmip6/)).
* chelsa-monthly: monthly bioclim variables from CHELSA #TO COMPLETE
* esacci-lc: ESA land cover from 1992 to 2020 (yearly) (see [ESA-CCI project](https://www.esa-landcover-cci.org/)).
* ghmts: global human modification index, a continuous 0-1 metric that reflects the proportion of a landscape modified, based on modeling the physical extents of 13 anthropogenic stressors (see [GHM project](https://sedac.ciesin.columbia.edu/data/set/lulc-human-modification-terrestrial-systems)).

```{r}
stac("http://io.biodiversite-quebec.ca/stac/") %>%
  collections() %>%
  get_request()

``` 

### Loading observations

Let's download [Glyptemys insculpta](https://en.wikipedia.org/wiki/Wood_turtle) observations from GBIF.

```{r}
obs <- rgbif::occ_data(scientificName = "Glyptemys insculpta", hasCoordinate = TRUE, limit = 1000)
obs <- dplyr::select(obs$data, decimalLongitude, decimalLatitude) %>%
  dplyr::rename(lon = decimalLongitude) %>%
  dplyr::rename(lat = decimalLatitude) 
head(obs)
```

We reproject the decimal longitude and latitude to a user-specified projection system. This conversion to a projected system is essential to interact with the cube.

```{r}
srs.obs <-"EPSG:4326" # initial observations projection system
srs.cube <- "EPSG:6623" # targeted projected projection system
obs.coords.proj <- create_projection(obs, lon = "lon", lat = "lat", 
                                       srs.obs, srs.cube)

```

## Loading current climate data

We load the cube from the collections "chelsa-climate",  using the observations points as an extent (option  use.obs = T and adding the observations points in parameter obs). If use.obs is FALSE, it is necessary to manually fill the args left, right, top and bottom. We subset the layers to three variables (to select all the variables, set the argument layers to NULL).
CHELSA current climate correspond to the aggregation of climate data from 1981 to 2010. Thus, here, t0 = t1 (by default, initial date of the period). 
The temporal resolution corresponds to a duration, which has to be in the [ISO 8601] format (https://en.wikipedia.org/wiki/ISO_8601), e.g. "P1Y" for a yearly aggregation, "P1M" for a monthly aggregation. Here it is not a useful parameter as data are yet aggregated over the 30-years period. 

Spatial resolution is set to 1 km.

Aggregation and resampling methods can be chosen, corresponding to the function applied to the pixel values when aggregating or disaggregating.


```{r, warning=FALSE}
get_info_collection(stac_path =
                                  "http://io.biodiversite-quebec.ca/stac/",
                                limit = 5000,
                                collections = c('chelsa-clim'),
                                bbox = NULL)

```

```{r, warning=FALSE}
layers <- c("bio6", "bio13")
obs_data <- read_delim("C:/GitHub/biab-2.0/output/loadGbif /loadGbif_R/8a3683f3d98459637d4cdf5e6a02521c/obs_data.tsv", 
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE)
proj = "EPSG:6623"
obs.coords.proj <- create_projection(obs_data, lon = "decimalLongitude", lat = "decimalLatitude", 
                              proj_from = "+proj=longlat +datum=WGS84",
                              proj_to = proj,new_lon = "lon", new_lat = "lat") %>% dplyr::select(lon, lat)

cube <- 
  load_cube(stac_path = "http://io.biodiversite-quebec.ca/stac/",
           limit = 5000, 
           collections = c("chelsa-clim"), 
           use.obs = T,
           obs = obs.coords.proj,
           buffer.box = 0,
          layers = layers,
           srs.cube = srs.cube,
           t0 = "1981-01-01",
           t1 = "1981-01-01",
           spatial.res = 1000, # in meters
           temporal.res = "P1Y",
           aggregation = "mean",
           resampling = "near") 
```
```{r, warning=FALSE}
cube %>% select_bands(c("bio6", "bio13")) %>%  plot()

```

The cube can be converted to classic raster format in R. The custom-built cube_to_raster function allows converting either to the raster or terra format, using the argument format.

```{r, warning=FALSE}
my.raster <- cube_to_raster(cube, format = "raster")
my.raster
```

To extract the values of variables for a set of points, we use the function extract_geom from the gdalcubes package. Points have to be in a sf format.
```{r, warning=FALSE}
value.points <- gdalcubes::extract_geom(cube, sf::st_as_sf(obs.coords.proj, coords = c("lon", "lat"),
                                                         crs = srs.cube)) 

obs.values <- dplyr::bind_cols(dplyr::select(value.points, FID, time) %>% dplyr::rename(id = FID),
          data.frame(obs.coords.proj),
          dplyr::select(value.points, dplyr::all_of(layers)))

head(obs.values)
```

## Loading future climate data

We load the climate data projection from the collection "chelsa-clim-proj", still using the observations points as an extent. The structure of the function is very similar to load_cube, except that the user has to select the climatic scenario (for now, "ssp126", "ssp370", or "ssp585") and the time span ("2041-2070" or "2071-2100").


```{r, warning=FALSE}
cube.future <- load_cube_projection(stac_path = 
                                    "http://io.biodiversite-quebec.ca/stac/",
                                  limit = 5000, 
                                  collections = c('chelsa-clim-proj'), 
                                  use.obs = T,
                                  obs = obs.coords.proj,
                                  buffer.box = 0,
                                  rcp = 'ssp126', #ssp126, ssp370, ssp585
                                  bbox = NULL, #layers = layers,
                                  srs.cube = srs.cube, 
                                  variable = "bio1",
                                  time.span = "2041-2070", #2041-2070 or 2071-2100
                                  spatial.res = 1000, 
                                  temporal.res  = "P1Y", 
                                  aggregation = "mean",
                                  resampling = "near") 
```

Variable extraction also uses the extract_geom function but the output is different. The output table contains one column
per variable and per climatic model. 
```{r, warning=FALSE}
value.points.future <- extract_geom(cube.future, sf::st_as_sf(obs.coords.proj, coords = c("lon", "lat"),
                                                         crs = srs.cube))
value.points.future <- dplyr::bind_cols(data.frame(obs.coords.proj),
                              value.points.future) %>% dplyr::rename(id = FID)

head(value.points.future)
```

We extract the name of the variable, the year, the model and the scenario from the column names, then we aggregate the value by point and bioclimatic variable using the mean (we could use another function such as mdeian, min, etc.)
```{r, warning=FALSE}
value.points.future <- value.points.future %>% tidyr::pivot_longer(
    cols = dplyr::starts_with("bio"),
    names_to = c("variable", "year", "model", "scenario"), 
    names_pattern = "(.*)_(.*)_(.*)_(.*)",
    values_to = "value"
  )  %>%
  dplyr::group_by(id, lon, lat, variable) %>% dplyr::summarise(mean = mean(value))
head(value.points.future)

```

We eventually convert the data into a tidy dataframe.

```{r, warning=FALSE}
value.points.future <- value.points.future %>%
     tidyr::pivot_wider(names_from = variable, values_from = mean)
head(value.points.future)

```
## Loading land cover data

```{r, warning=FALSE}
get_info_collection(stac_path =
                                  "http://io.biodiversite-quebec.ca/stac/",
                                limit = 5000,
                                collections = c('esacci-lc'),
                                bbox = NULL)

```

```{r, warning=FALSE}
cube <- 
  load_cube(stac_path = "http://io.biodiversite-quebec.ca/stac/",
           limit = 5000, 
           collections = c("esacci-lc"), 
           use.obs = T,
           obs = obs.coords.proj,
           buffer.box = 0,
           srs.cube = srs.cube,
           t0 = "2000-01-01",
           t1 = "2000-12-31",
           spatial.res = 1000, # in meters
           temporal.res =  "P1Y",
           aggregation = "mean",
           resampling = "near") 

```

```{r, warning=FALSE}
plot(cube %>% select_bands("esacci-lc-2000"))
```

## Loading global human modification index

```{r, warning=FALSE}
cube <- 
  load_cube(stac_path = "http://io.biodiversite-quebec.ca/stac/",
           limit = 5000, 
           collections = c("ghmts"), 
           use.obs = T,
           obs = obs.coords.proj,
           buffer.box = 0,
           srs.cube = srs.cube,
           t0 = "2016-01-01",
           t1 = "2016-01-01",
           spatial.res = 1000, # in meters
           temporal.res = "P1Y",
           aggregation = "mean",
           resampling = "near") 

```

## Loading monthly climate data
```{r, warning=FALSE}
get_info_collection(stac_path =
                                  "http://io.biodiversite-quebec.ca/stac/",
                                collections = c('chelsa-monthly'),
                                bbox = NULL)

```

```{r, warning=FALSE}
cube <- 
  load_cube(stac_path = "http://io.biodiversite-quebec.ca/stac/",
           limit = 5000, 
           collections = c("chelsa-monthly"), 
           use.obs = T,
           obs = obs.coords.proj,
           buffer.box = 0,
           srs.cube = srs.cube,
           t0 = "2016-01-01",
           t1 = "2016-03-01",
           variable = "tasmax",
           spatial.res = 1000, # in meters
           temporal.res = "P1Y",
           aggregation = "mean",
           resampling = "near") 


```

```{r, warning=FALSE}
plot(cube)
```
```{r, warning=FALSE}
get_info_collection(stac_path =
                             "https://planetarycomputer.microsoft.com/api/stac/v1/",
                            limit = NULL,
                            collections = c('cop-dem-glo-90'),
                            bbox = NULL)
```                            
```{r, warning=FALSE}
#layers <- c("bio6", "bio13")

cube <- 
  load_cube(stac_path = "https://planetarycomputer.microsoft.com/api/stac/v1/",
           limit = NULL, 
           collections = c("cop-dem-glo-90"), 
           use.obs = F,
           obs = obs.coords.proj,
           buffer.box = 0,
          bbox = data.frame("xmin" = -248000, "xmax" = -157719, "ymin" = 344267, "ymax" = 500000),
           srs.cube = srs.cube,
           t0 = "2021-04-22",
           t1 = "2021-04-22",
           spatial.res = 1000, # in meters
           temporal.res = "P1Y",
           aggregation = "mean",
           resampling = "near") 
```
```{r, warning=FALSE}
get_info_collection(stac_path =
                                  "http://io.biodiversite-quebec.ca/stac/",
                                collections = c('soilgrids'),
                                bbox = NULL)

cube <- 
  load_cube(stac_path = "http://io.biodiversite-quebec.ca/stac/",
           limit = 5000, 
           collections = c("soilgrids"), 
           use.obs = T,
           obs = obs.coords.proj,
           buffer.box = 0,
          layers = "sand_0-5cm",
          t0 = "2016-07-04T00:00:00Z",
          t1= "2016-07-04T00:00:00Z",
           srs.cube = srs.cube,
           spatial.res = 1000, # in meters
           temporal.res = "P1Y",
           aggregation = "mean",
           resampling = "near") 
```

### To continue: usefull ressources

* gdalcube doc: [https://cran.r-project.org/web/packages/gdalcubes/gdalcubes.pdf](https://cran.r-project.org/web/packages/gdalcubes/gdalcubes.pdf)
* Examples based on gdalcubes library: [https://github.com/appelmar/gdalcubes_R](https://github.com/appelmar/gdalcubes_R)
